Simulations of Ballot Polling Risk-Limiting Audits

### 2023-02-13 providence/sections/intro.tex
not clear enough
some changes to intro

A tabulation audit will either end with a declaration that an election outcome is correct, or escalate to a full hand count. The integrity of an audit may be judged by how it deals with incorrect outcomes. A {\em risk-limiting audit (RLA)} guarantees a minimum probability that it will perform as it is supposed to (escalate to a full hand count) if the outcome is incorrect. Equivalently, this is also a guarantee limiting the probability of audit error when the outcome is incorrect. The {\em risk limit} of an RLA is the (guaranteed) maximum probability that an incorrect election outcome would be declared correct (and not proceed to a full hand count). Lower risk limits are better.

@@

\subsection{Background on RLAs}
All RLAs sample one or more ballots at random; we will refer to each such set of ballots as a {\em round}. The ballots are manually examined, and a stopping condition computed, which determines whether (a) the audit ends in success (the election outcome is declared correct) or (b) another round should be drawn (more information is needed before a determination). In principle, the stopping condition could indicate a third option too: (c) the audit proceeds to a full manual hand count, which is generally not incorporated \cite{bravo}. A manual hand count presents significant logistical challenges, and there is always a chance that it will have been unnecessary. Election officials would typically decide to perform a full manual hand count if the audit does not stop in spite of drawing a large number of ballots, typically over multiple rounds. They would be influenced by the certification deadline, the estimated number of human hours required for another round, the logistical costs of a full hand count, and the impact of any decision on citizen confidence.

There are three types of RLAs: ballot comparison RLAs, batch comparison RLAs and ballot polling RLAs. This paper focuses on ballot polling RLAs which have been used in a number of US state pilots (California, Georgia, Indiana, Michigan, Ohio, Pennsylvania, Virginia and elsewhere) and in real statewide audits (Georgia, Virgina) \cite{vv_audits} as well as in audits of smaller jurisdictions, such as Montgomery County, Ohio \cite{usenix_minerva}. Ballot and batch comparison RLAs are described in \ref{sec:related}.

@@

Ballot polling audits proceed as follows. 
\begin{enumerate}
\item A first round \cite{usenix_minerva} size---the number of ballots sampled before first checking the stopping condition---is chosen. 
\item Ballots on the ballot manifest are sampled uniformly at random using a pseudorandom number generator typically seeded by a natural source of randomness like rolling dice.
\item The physical ballots are found and manually interpreted, recording the manual interpretations. 
\item Based on the manual interpretations, the stopping condition is computed. 
\item If more ballots are to be drawn, the next round size is chosen, and we go back to step 2. 
\end{enumerate}

@@

@@

Election officials typically draw ballots in large round sizes, see for example \cite{va-2022,RI-report}. Note also that, in addition to allowing users to directly enter a round size or choose the expected number of ballots drawn by \BRAVO, Arlo provides choices of stopping probabilities of $0.9$, $0.8$ and $0.7$ (\cite{usenix_minerva} provides estimates of these round sizes for each state in the 2020 US Presidential election, for a $0.9$ probability of stopping; some numbers are in tens and hundreds of thousands of ballots, and thousands of ballots are quite common). Further, at both audits we attended, election officials chose stopping probabilities of $0.9$ and we are not aware of any ballot polling RLA performed on ballots cast in a governmental election that drew ballots one at a time (though the stopping condition can be computed one ballot at a time, the ballots are drawn in rounds). \BRAVO hence cannot be used as a B2 audit in these scenarios.

For use as an R2 audit, the \BRAVO stopping condition can be applied once at the end of each round (End-of Round (EoR)), or retroactively after each ballot drawn if ballot order is retained (Selection-Ordered (SO)). SO \BRAVO is closer to the original B2 \BRAVO, and requires fewer ballots on average than EoR \BRAVO. But it requires the additional effort of tracking the order of ballots.

@@

Referring to the ballot polling audit steps described in section \ref{sec:polling} above, we further assume that a secure PRNG is used; the seed is generated uniformly at random in a public process; the process of locating ballots is publicly observable and the located ballots can be viewed by the public. Because the PRNG is well-defined, as is the stopping condition, we may assume that the stopping condition is correctly computed, because it can be checked by the public through knowledge of the seed and the drawn ballots. Thus the only variable is round size.

@@

\subsection{Open Questions}
Zag\'{o}rski {\em et al.} propose ballot polling RLA \Minerva \cite{usenix_minerva}, which does not need ballot order and relies only on sample and round tallies. They prove that it is risk-limiting when the number of relevant ballots drawn in each round is pre-determined before any ballots are examined; that is, for a weak adversary. They do not address the case of a strong adversary (such as an audit insider) who can determine the size of the next round after knowing what votes are on the ballots sampled thus far. In particular, an open question about \Minerva is whether the computation of a risk limit assuming a weak adversary applies to an attack by a strong adversary, or is the risk limit computation incorrect when the adversary is strong? Can the strong adversary increase the audit's error probability beyond its declared risk limit? Or is there no probabilistic adversarial advantage to being able to compute next round sizes after knowing the drawn sample? We do not answer this question, and to our knowledge, it remains open.

@@

A major limitation of our understanding of the ballot polling problem as a community is that we use the number of ballots drawn or values proportional to this number \cite{mclaughlin_thesis,bernhard-diss,RI-report} as measures of the workload of an audit. If this were a correct measure of the workload of an audit, we would want to use B2 audits (round size is one) and make decisions about stopping the audit after drawing each ballot, because this leads to the smallest expected number of ballots. As described above, election officials, on the other hand, greatly prefer drawing many ballots at once. From conversations with election officials and staff members of Verified Voting, Brennan Center and Common Cause who have been training election officials to perform RLAs, we estimate that this preference is likely due to the following. 
\begin{description}
\item Firstly, each round has an overhead workload as well, including setting up the round and communicating among the various localities involved in conducting the audit (for example, audits of statewide contests involve the drawing of ballots at county offices where the ballots are stored). 
\item Secondly, there is an overhead to finding a storage box and unsealing it. For large round sizes, multiple ballots may be drawn at once from a box, and the number of boxes retrieved is smaller than the number of ballots (storage boxes commonly contain many hundreds of ballots each). For smaller round sizes, the number of times a box is retrieved would be roughly identical to the number of ballots drawn, as it is unlikely that a single box will hold multiple ballots from the sample. 
\item Finally, in the current environment of misinformation, election officials fear a misleading audit sample (falsely indicating that the loser won), preferring to structure audits to reduce the chances of such samples, thus implicitly choosing larger round sizes. 
\end{description}
Thus the workload of an audit is not simply a linear (or affine) function of the number of ballots drawn. Relatedly, an optimal round schedule is not completely determined by the expected number of ballots drawn. It depends on other variables as well. The consideration of all these variables is necessary while planning an audit.

\subsection{Our Contributions}
Our primary contribution is a new RLA, \Providence, which gives the efficiency of \Minerva and is also resistant to a strong adversary. The stopping condition for \Minerva does not take into account the sample obtained in previous rounds, and, in \cite{usenix_minerva}, its risk limit is estimated through weighted averages across multiple rounds, assuming that round sizes do not depend on the previous sample. We are able to derive a new stopping condition for which a far simpler proof of the risk-limiting property is possible. In particular, this proof does not require an assumption about round sizes.

@@

@@

\subsection{Organization} 
Sections \ref{sec:model} and \ref{sec:related} provide background, describing the audit model and related work respectively. Section \ref{sec:prov} describes the \Providence audit, Section \ref{sec:sims} the simulations comparing the number of ballots drawn using various ballot polling audits and section \ref{sec:pilot} the use of \Providence in an audit carried out by the Board of Elections of Rhode Island. Section \ref{sec:workload} presents our workload model and describes its use for a ballot polling audit using details of the 2020 US Presidential election in the state of Virginia. Our conclusions, the availability of an audit implementation and acknowledgements may be found in sections \ref{sec:conc}, \ref{sec:avail} and \ref{sec:ack} respectively.

>>>>

A tabulation audit will either end with a declaration that an election outcome is correct, or escalate to a full hand count. The integrity of an audit may be judged by how it deals with incorrect outcomes. A {\em risk-limiting audit (RLA)} guarantees a minimum probability that it will perform as it is supposed to (escalate to a full hand count) if the outcome is incorrect. Equivalently, this is also a guarantee limiting the probability of audit error when the outcome is incorrect. The {\em risk limit} of an RLA is the (guaranteed) maximum probability that an incorrect election outcome would be declared correct. Lower risk limits are better.

@@

\subsection{Background on RLAs}
All RLAs sample one or more ballots at random; we will refer to each such set of ballots as a {\em round}. The ballots are manually examined, and a stopping condition computed, which determines whether (a) the audit ends in success (the election outcome is declared correct) or (b) another round should be drawn (more information is needed before a determination). In principle, the stopping condition could indicate a third option too: (c) the audit proceeds to a full manual hand count. This option is generally not incorporated \cite{bravo}. A manual hand count presents significant logistical challenges, and there is always a chance that it will have been unnecessary. Election officials would typically decide to perform a full manual hand count if the audit does not stop in spite of drawing a large number of ballots, typically over multiple rounds. They would be influenced by the certification deadline, the estimated number of human hours required for another round, the logistical costs of a full hand count, and the impact of any decision on citizen confidence.

There are three types of RLAs: ballot comparison RLAs, batch comparison RLAs and ballot polling RLAs. This paper focuses on ballot polling RLAs which have been used in a number of US state pilots (California, Georgia, Indiana, Michigan, Ohio, Pennsylvania, Virginia and elsewhere), real statewide audits (Georgia, Virgina) \cite{vv_audits} and  audits of smaller jurisdictions, such as Montgomery County, Ohio \cite{usenix_minerva}. Ballot and batch comparison RLAs are described in \ref{sec:related}.

@@

Ballot polling audits proceed as follows. 
\begin{enumerate}
\item The ballot manifest is published. 
\item A first round size \cite{usenix_minerva} is chosen. %\fpo{the sentence is hard to understand}. pv: fixed
\item Ballots on the ballot manifest are sampled uniformly at random using a pseudorandom number generator, which is typically seeded by a natural source of randomness like rolling dice.
\item The physical ballots are found and manually interpreted, recording the manual interpretations. 
\item Based on the manual interpretations, the stopping condition is computed. 
\item If more ballots are to be drawn, the next round size is chosen, and we go back to step 2. 
\end{enumerate}

@@

The stopping condition for \BRAVO, \Minerva and \Providence is described for a contest with two participants. Audits of multiple-candidate contests are treated as multiple audits of the pairwise contest between the announced winner and all other candidates. \BRAVO is also used for what are termed {\em stratified} audits [REF], where a single contest is audited using different approaches (ballot polling, ballot comparison, batch comparison) in different counties. When a ballot polling audit is used for the entire set of ballots, the stopping condition assumes a tied outcome (the hardest incorrect outcome to detect) when measuring the risk limit. In a stratified audit, however, when testing individual units such as counties, one may need to assume a different outcome. This is because, for example, if a county with an announced large number of ballots for the winner were actually tied, the overall outcome would be incorrect. The proof of \Minerva's risk-limiting property assumes a tied outcome while measuring risk, and it is not immediately obvious how the proof may be modified to allow other possibilities. Thus, while \Minerva may be used for multi-candidate contests when all counties participate in the ballot polling audit, it cannot be used for stratified audits. \BRAVO, and \Providence, on the other hand, can.

@@

Election officials typically draw ballots in large round sizes, see for example \cite{va-2022,RI-report}. Note also that, in addition to allowing users to directly enter a round size or choose the expected number of ballots drawn by \BRAVO, Arlo provides choices of stopping probabilities of $0.9$, $0.8$ and $0.7$ (\cite{usenix_minerva} provides estimates of these round sizes for each state in the 2020 US Presidential election, for a $0.9$ probability of stopping; some numbers are in tens and hundreds of thousands of ballots, and thousands of ballots are quite common). For the two audits we attended, election officials chose stopping probabilities of at least $0.9$. We are not aware of any ballot polling RLA performed on ballots cast in a governmental election that drew ballots one at a time (though the stopping condition can be computed one ballot at a time, the ballots are drawn in rounds). \BRAVO hence cannot be used as a B2 audit in these scenarios.

For use as an R2 audit, the \BRAVO stopping condition can be applied once at the end of each round (End-of Round (EoR)), or retroactively after each ballot drawn if ballot order is retained (Selection-Ordered (SO)). SO \BRAVO is closer to the original B2 \BRAVO, and requires fewer ballots on average than EoR \BRAVO. But it requires the additional effort of tracking the order of ballots, and should be expected to be inefficient because it does not use the information in the ballots drawn at the end.

@@

Referring to the ballot polling audit steps described in section \ref{sec:polling} above, we further assume that the ballot manifest is verified by the compliance audit; a secure PRNG is used; the seed is generated uniformly at random in a public process; the process of locating ballots is publicly observable and the located ballots can be viewed by the public. Because the PRNG is well-defined, as is the stopping condition, we may assume that the stopping condition is correctly computed, because it can be checked by the public through knowledge of the seed and the drawn ballots. Thus the only variable is round size.

@@

\subsection{Open Questions}
Zag\'{o}rski {\em et al.} propose ballot polling RLA \Minerva \cite{usenix_minerva}, which does not need ballot order and relies only on sample and round tallies. They prove that it is risk-limiting when the number of relevant ballots drawn in each round is pre-determined before any ballots are examined; that is, for a weak adversary. They do not address the case of a strong adversary (such as an audit insider) who can determine the size of the next round after knowing what votes are on the ballots sampled thus far. In particular, an open question about \Minerva is whether the computation of a risk limit assuming a weak adversary applies to an attack by a strong adversary, or is the risk limit computation incorrect when the adversary is strong? Can the strong adversary change the election outcome and then increase the audit's error probability beyond its declared risk limit? Or is there no probabilistic adversarial advantage to being able to compute next round sizes after knowing the drawn sample? We do not answer this question, and to our knowledge, it remains open.

@@

A major limitation of our understanding of the ballot polling problem as a community is that we use the number of ballots drawn or values proportional to this number \cite{mclaughlin_thesis,bernhard-diss,RI-report} as measures of the workload of an audit. If this were a correct measure of the workload of an audit, we would want to use B2 audits (round size is one) and make decisions about stopping the audit after drawing each ballot, because this leads to the smallest expected number of ballots. As described above, election officials, on the other hand, greatly prefer drawing many ballots at once. From conversations with election officials and staff members of Verified Voting, Brennan Center and Common Cause who have been training election officials to perform RLAs, we estimate that this preference is likely due to the following. 
\begin{description}
\item Firstly, each round has an overhead workload as well, including setting up the round and communicating among the various localities involved in conducting the audit (for example, audits of statewide contests involve the drawing of ballots at county offices where the ballots are stored). 
\item Secondly, there is an overhead to finding a storage box and unsealing it. For large round sizes, multiple ballots may be drawn at once from a box, and the number of boxes retrieved is smaller than the number of ballots (storage boxes commonly contain many hundreds of ballots each). 
%For smaller round sizes, the number of times a box is retrieved would be roughly identical to the number of ballots drawn, as it is unlikely that a single box will hold multiple ballots from the sample. 
\item Finally, in the current environment of misinformation, election officials fear a misleading audit sample (falsely indicating that the loser won), preferring to structure audits to reduce the chances of such samples, thus implicitly choosing larger round sizes. 
\end{description}
Thus the workload of an audit is not simply a linear (or affine) function of the number of ballots drawn. Relatedly, an optimal round schedule is not completely determined by the expected number of ballots drawn. It depends on other variables as well. The consideration of all these variables is necessary while planning an audit.

\subsection{Our Contributions}
Our primary contribution is a new RLA, \Providence, which gives the efficiency of \Minerva and is also resistant to a strong adversary. Additionally, \Providence does not assume that the null hypothesis is a tied election and hence is more broadly useful than \Minerva.

The stopping condition for \Minerva does not take into account the sample obtained in previous rounds, and, in \cite{usenix_minerva}, its risk limit is estimated through weighted averages across multiple rounds, assuming that round sizes do not depend on the previous sample. Further, it assumes that the null hypothesis is a tie. A far simpler proof is possible for the risk-limiting property of \Providence. In particular, the proof does not require an assumption about round sizes and does not assume any particular fractional vote distribution for the null.

@@

\Providence may be used in any audit, stratified or otherwise, where audited contests may be expressed as pairwise plurality contests: for example, it can be used in plurality and majoritarian elections.

@@

\subsection{Organization} 
Sections \ref{sec:model} and \ref{sec:related} provide background, describing the audit model and related work respectively. Section \ref{sec:prov} describes the \Providence audit, section \ref{sec:sims} the simulations comparing the number of ballots drawn using various ballot polling audits and section \ref{sec:pilot} the use of \Providence in an audit carried out by the Board of Elections of Rhode Island. Section \ref{sec:workload} presents our workload model and describes its use for a ballot polling audit using details of the 2020 US Presidential election in the state of Virginia. Our conclusions, the availability of an audit implementation and acknowledgements may be found in sections \ref{sec:conc}, \ref{sec:avail} and \ref{sec:ack} respectively.

### 2023-02-13 providence/sections/related.tex
some changes to intro

For completeness, we describe other types of RLAs below. 
\begin{description}
\item In a ballot comparison RLA \cite{RLA}, the manual interpretation of each sampled ballot is compared to the corresponding Cast Vote Record (CVR), which is the machine interpretation of the ballot. Ballot comparison RLAs require the fewest ballots of all known RLA approaches, but also require a means of identifying the CVR corresponding to a particular ballot. A typical approach is to use a ballot serial number on both paper ballot and CVR. When voters vote in precincts, however, serial numbers on ballots can enable the correlation of ballots with voters, and ballots are typically not numbered. Additionally, some voting systems do not record a CVR. One may perform a transitive audit by rescanning unnumbered voted ballots with special scanners which produce CVRs and also print numbers on the ballots as they are scanned. This requires an investment in a sufficiently large number of such printers and in the human effort of rescanning ballots, and is not always feasible. 
\item A batch comparison RLA \cite{RI-report} samples batches of ballots (typically, a batch is a storage box of ballots) and compares the manual tally of each sampled batch with the announced tally of that batch. Thus, while this type of audit does not need CVRs, it does need both a ballot manifest and a public declaration of the tally of each batch. This approach typically requires the sampling of a very large number of ballots. However, the process is most similar to that which election officials already use when they perform fixed-percentage post-election audits, where, for example, 2\% of the batches are manually tallied. (Note that, for an RLA, the number of batches tallied is not fixed; the risk limit is. A smaller number of batches might be sufficient, or a larger number necessary, for an audit with the required risk limit.)
\end{description}

>>>>

For completeness, we describe other types of RLAs below.  
\begin{description}
\item In a ballot comparison RLA \cite{RLA}, the manual interpretation of each sampled ballot is compared to the corresponding Cast Vote Record (CVR), which is the machine interpretation of the ballot. Ballot comparison RLAs require the fewest ballots of all known RLA approaches, but also require a means of identifying the CVR corresponding to a particular ballot. A typical approach is to use a ballot serial number on both paper ballot and CVR. When voters vote in precincts, however, serial numbers on ballots can enable the correlation of ballots with voters, and ballots are typically not numbered. Additionally, some voting systems do not record a CVR. One may perform a transitive audit by rescanning unnumbered voted ballots with special scanners which produce CVRs and also print numbers on the ballots as they are scanned. This requires an investment in a sufficiently large number of such printers and in the human effort of rescanning ballots, and is not always feasible. There is recent work on {\em adaptive comparison} RLAs, which relies on scanning only the batches necessary \cite{adaptive-comparison}. In either case, the workload of a transitive ballot comparison audit, measured in hours, would not be simply a linear function the number of ballots drawn, but would include a term which would depend on the total number of ballots rescanned. 
\item A batch comparison RLA \cite{RI-report} samples batches of ballots (typically, a batch is a storage box of ballots) and compares the manual tally of each sampled batch with the announced tally of that batch. Thus, while this type of audit does not need CVRs, it does need the publication of the tally of each batch. Batch comparison typically requires the sampling of a very large number of ballots, larger than polling audits except for small enough margins. However, the process is most similar to that which election officials already use when they perform fixed-percentage post-election audits, where, for example, 2\% of the batches are manually tallied. (Note that, for an RLA, the number of batches tallied is not fixed; the risk limit is. A smaller number of batches might be sufficient, or a larger number necessary, for an audit with the required risk limit.)
\end{description}
